Gaussian Process regression for astronomical time-series

### 2022-11-09 src/tex/ms.tex
[showyourwork] overleaf sync

The kernel function can be any positive scalar function that gives rise to a positive semi-definite covariance matrix over the input domain, but some will be more useful than others.
Given this large decision space, you may be wondering how to choose the right kernel function for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel functions have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that would naturally motivate their use in certain circumstances.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can recast as a GP model with a specific kernel function.
% \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

Even if you don't have a formal physics-based justification for your model, you may be able to identify the key scales of your problem and design a kernel function that captures these features.
The usual approach to this problem is to take sums and products of commonly used kernel functions to select a set of  models that have the needed covariance structure, and then combine or select between these choices.
For example, if there is expected to only be a single non-periodic timescale in the problem of interest, you could list all the two-parameter non-periodic kernel functions and use a numerical model selection technique as described below.

@@

Another example that is commonly used in the astrophysics literature \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H} is the following ``quasi-periodic'' kernel function
\begin{equation}
\label{eq:kqp}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,{\lengthscale_1}^2} -\Gamma\,\sin^2\left[\frac{\pi\,\dt}{\lengthscale_2}\right] \right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale_1,\,\lengthscale_2,\,\gamma\}$, which has a period of $\lengthscale_2$, and a decoherence timescale of $\lengthscale_1$. The hyperparameter $\Gamma$ controls the extent to which the periodic component of the signal resembles a simple sinusoid or is more complex.

@@

Similarly, integration or convolution also have similar properties.
For example, real time series observations are made with finite exposure time $\Delta$, and therefore, if the underlying stochastic process $f(t)$ is modelled as a GP with covariance $k(t_i,t_j;\hyperparams)$, the observations are actually of
\begin{equation}
  f_\mathrm{int}(t) = \int_{t-\Delta/2}^{t+\Delta/2} f(t^\prime)\,\mathrm{d}t^\prime \quad,
\end{equation}
which has a covariance of
\begin{equation}
  k_\mathrm{int}(t_i,t_j;\hyperparams) = \int_{t_i-\Delta/2}^{t_i+\Delta/2}\int_{t_j-\Delta/2}^{t_j+\Delta/2} k(t,t^\prime;\hyperparams)\,\mathrm{d}t\,\mathrm{d}t^\prime \quad,
\end{equation}
which can be evaluated in closed form for some standard kernel functions.
Another context where integrated GPs have recently been used for astrophysics (albeit not for time series), is as a model of the Galactic dust distribution \citep{2022arXiv220206797M}.

@@

GP models were first applied to the simulated observations of the Sun's total irradiance variations by \citet{2012MNRAS.419.3147A}, to test a new method to simulate Radial Velocity (RV) variations based space-based photometry on solar data. The context of this work was the large numbers of candidate transiting exoplanets being discovered at the time by the \textit{CoRoT} and \textit{Kepler} space missions. Detecting the planets signals in RV was needed to confirm the planetary nature of the candidates by measuring their masses, but was being hampered by the apparent RV variations caused by active regions. Using simple geometric considerations, \citet{2012MNRAS.419.3147A} derived a simple (approximate) relationship between the flux perturbation caused by active regions, $F$, and their RV signature, which depends on both $F$ and its time-derivative $F'$. They tested this data-driven ``$FF'$ method'' on simulated photometric and RV observations of the Sun-as-a-star produced using resolved magnetograms, using a GP as a principled smoothing tool to evaluate $F$ and $F'$ from the photometry, and comparing the results to the simulated RVs. \citet{2012MNRAS.419.3147A}  tested a number of covariance functions, both aperiodic and Quasi-Periodic (QP), and found the former to be preferred in the solar case, as expected if most active regions on the Sun do not persist for much longer than its rotation period. More active and/or more rapidly rotating stars display variability that is coherent over multiple rotation periods, making QP GPs the model of choice. For example, \citet{2014MNRAS.443.2517H} used a QP kernel when applying the aforementioned $FF'$ method to the active planet-host star CoRoT-7.

Today, using GPs to model light curves containing stellar variability is standard practice. The kernel most frequently used for this purpose is the aforementioned QP kernel, whose simplicity and flexibility make it a popular choice. \citet{2018MNRAS.474.2094A} implemented a Bayesian inference framework based on this kernel to measure accurate rotation periods from \textit{Kepler} light curves. The very complex dependence of the likelihood surface on the parameters, particularly the period, makes this challenging, and careful tuning of the posterior sampling strategy is required.

@@

\subsubsection{Stellar activity in Radial Velocities} \label{sec:RVact}
Given the remarkable instrumental precision achieved by modern RV spectrographs, stellar activity is nowadays the key factor limiting the sensitivity of RV surveys low-amplitude and/or long-period planet signals. The effect of active regions on RV observations is two-fold. First, as dark spots rotate on the stellar surface, they distort the profile of spectral lines, removing a small contribution first from the red wing of each spectral line, then from the line core, and then from the blue wing. The second effect is more subtle: in the absence of active regions, spectral lines of Sun-like stars display a net blue-shift due to granulation: the emission from the hot, up-welling material in the granules, which is blue-shifted, dominates over that from the cooler material falling back down in the inter-granular lanes). In a facula, i.e.\ a region of enhanced magnetic flux density compared to the ``clean'' photosphere, convection is suppressed, leading to a localised reduction in this ``convective blue-shift''. Faculae have very small photometric contrast but can cover a much larger area than dark spots, so their RV signature can dominate over that of spots, especially for moderately slow rotators like the Sun with modest magnetic fields \citep{2010A&A...512A..39M}.

@@

\begin{figure}[ht]
  \centering
  \script{scaling.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/scaling.pdf}
  \caption{How the computational cost of a single evaluation of the GP likelihood function (\autoref{eq:gp_like}) scales with the size of the dataset using different algorithms.
  In each case, the kernel function is a one-dimensional Mat\'ern-3/2 with randomly generated data.
  The top two lines labelled ``direct (CPU)'' and ``direct (GPU)'' show the cost of directly evaluating \autoref{eq:gp_like} directly using the Cholesky factorization of $\mathbf{K}$ provided by a tuned and optimised linear algebra library.
  The CPU solver was run using a single-threaded implementation, and the GPU version was executed on an \textsf{A100 NVIDIA} GPU.
  In both cases, the asymptotic scaling of the algorithm was approximately $\mathcal{O}(N^3)$ for $N$ data points.
  The three lines labelled ``celerite ($J$ terms)'' show the performance of the \project{celerite} algorithm \citep[][as implemented by the \project{tinygp} package]{celerite} for a sum of $J$ Mat\'ern-3/2 kernels, since the computational cost of that method scales with the ``complexity'' of the kernel.
  The asymptotic performance of each of these curves is linear in the number of data points, $\mathcal{O}(N)$.}
  \label{fig:scaling}
\end{figure}

@@

Alongside this review, we have released all the source code to generate the figures in this document using the \project{tinygp} library\footnote{\url{https://tinygp.readthedocs.io}} \citep{tinygp}, which is a relatively new open source \project{Python} library for GP modelling.
While this library hasn't yet been widely used in the astrophysics literature, we believe\footnote{Note that the development of the \project{tinygp} library is led by D.\ Foreman-Mackey, an author of this review.} that it is a good choice for many applications.
\project{tinygp} uses the \project{JAX} library as its computational backend, giving it good computational performance (including support for GPU hardware) with a pure-\project{Python} user interface.
\project{JAX} also provides support for automatic differentiation \citep[e.g.,][]{autodiff}, which can be used to accelerate all the inference schemes described in Sections~\ref{sec:inference} and~\ref{sec:model_selection}.

@@

After a very brief historical introduction, we started this review by introducing two simulated examples, which demonstrate the two main classes of use for GPR in astronomy, namely to account for a nuisance signal (in the case of a planetary transit) or to model a stochastic signal of interest (in the quasar time-delay example). The source code for these and other worked examples in the review is available alongside this review in the form of {\sc Jupyter} notebooks that can be accessed by clicking on the icon next to each figure caption, and interested users are encouraged to use these to further their understanding of the methodology and potentially adapt the code to their own datasets. Next, we gave a succinct introduction to the key equations underlying GPR regression, and sketched out a typical workflow for applying GPR to a new dataset. Like any data analysis method, GPR requires the user to make a number of choices, the most important of which is the choice of covariance function. We listed some of the most widely used covariance functions, outlined how to construct more complex ones, and gave practical advice on how to assess the validity of one's modelling choices. These strategies include standard Bayesian model comparison but also cross-validation, which can be more appropriate in cases where predictive power is important.

@@

SA gave a set of lectures on GPR for time-domain astronomy as part of the 2021 \emph{Saas Fee Advanced Course in Astrophysics}, for which PDF lecture notes are available at \url{shorturl.at/hrstz}, along with {\sc Python} exercises in the form of {\sc Jupyter} notebooks. \suz{This simple URL reference will need to be replaced with a citation to the official lecture notes published by Springer if those are published before the present review.}

>>>>

The kernel function can be any positive scalar function that gives rise to a positive semi-definite covariance matrix over the input domain, but some will be more useful than others.
Given this large decision space, you may be wondering how to choose the right kernel function for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel functions have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that would naturally motivate their use in certain circumstances.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can be recast as a GP model with a specific kernel function.
% \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

Even if you don't have a formal physics-based justification for your model, you may be able to identify the key scales of your problem and design a kernel function that captures these features.
The usual approach to this problem is to take sums and products of commonly used kernel functions to construct a set of models that have the desired covariance structure, and then combine or select between these choices.
For example, if the problem of interest is expected to involve only a single non-periodic timescale, you could list all the two-parameter non-periodic kernel functions and use a numerical model selection technique as described below.

@@

Another example that is commonly used in the astrophysics literature \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H} is the following ``quasi-periodic'' kernel function
\begin{equation}
\label{eq:kqp}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,{\lengthscale_1}^2} -\Gamma\,\sin^2\left[\frac{\pi\,\dt}{\lengthscale_2}\right] \right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale_1,\,\lengthscale_2,\,\Gamma\}$, which has a period of $\lengthscale_2$, and a decoherence timescale of $\lengthscale_1$. The hyperparameter $\Gamma$ controls the extent to which the periodic component of the signal resembles a simple sinusoid or is more complex.

@@

Integration or convolution also have similar properties.
For example, real time series observations are made with finite exposure time $\Delta$, and therefore, if the underlying stochastic process $f(t)$ is modelled as a GP with covariance $k(t_i,t_j;\hyperparams)$, the observations are actually of
\begin{equation}
  f_\mathrm{int}(t) = \int_{t-\Delta/2}^{t+\Delta/2} f(t^\prime)\,\mathrm{d}t^\prime \quad,
\end{equation}
which has a covariance of
\begin{equation}
  k_\mathrm{int}(t_i,t_j;\hyperparams) = \int_{t_i-\Delta/2}^{t_i+\Delta/2}\int_{t_j-\Delta/2}^{t_j+\Delta/2} k(t,t^\prime;\hyperparams)\,\mathrm{d}t\,\mathrm{d}t^\prime \quad,
\end{equation}
which can be evaluated in closed form for some standard kernel functions.
Another context where integrated GPs have recently been used for astrophysics (albeit not for time series), is as a model of the Galactic dust distribution \citep{2022arXiv220206797M}.

@@

GP models were first applied to the simulated observations of the Sun's total irradiance variations by \citet{2012MNRAS.419.3147A}, to test a new method to simulate Radial Velocity (RV) variations based space-based photometry on solar data. The context of this work was the large numbers of candidate transiting exoplanets being discovered at the time by the \textit{CoRoT} and \textit{Kepler} space missions. Detecting the planet signals in RV was needed to confirm the planetary nature of the candidates by measuring their masses, but was being hampered by the apparent RV variations caused by active regions. Using simple geometric considerations, \citet{2012MNRAS.419.3147A} derived a simple (approximate) relationship between the flux perturbation caused by active regions, $F$, and their RV signature, which depends on both $F$ and its time-derivative $F'$. They tested this data-driven ``$FF'$ method'' on simulated photometric and RV observations of the Sun-as-a-star produced using resolved magnetograms, using a GP as a principled smoothing tool to evaluate $F$ and $F'$ from the photometry, and comparing the results to the simulated RVs. \citet{2012MNRAS.419.3147A}  tested a number of covariance functions, both aperiodic and Quasi-Periodic (QP), and found the former to be preferred in the solar case, as expected if most active regions on the Sun do not persist for much longer than its rotation period. More active and/or more rapidly rotating stars display variability that is coherent over multiple rotation periods, making QP GPs the model of choice. For example, \citet{2014MNRAS.443.2517H} used a QP kernel when applying the aforementioned $FF'$ method to the active planet-host star CoRoT-7.

Today, using GPs to model light curves containing stellar variability is standard practice. The kernel most frequently used for this purpose is the aforementioned QP kernel, whose simplicity and flexibility make it a popular choice. \citet{2018MNRAS.474.2094A} implemented a Bayesian inference framework based on this kernel to measure accurate rotation periods from \textit{Kepler} light curves. The very complex dependence of the likelihood surface on the parameters, particularly the period, makes this challenging, and careful tuning of the posterior sampling strategy is required.
These methods have since been adapted and applied to ground-based photometric surveys \citep{2020MNRAS.492.1008G}, and the shorter baseline \textit{K2} light curves \citep{2021ApJ...913...70G}.

@@

\subsubsection{Stellar activity in Radial Velocities} \label{sec:RVact}
Given the remarkable instrumental precision achieved by modern RV spectrographs, stellar activity is nowadays the key factor limiting the sensitivity of RV surveys to low-amplitude and/or long-period planet signals. The effect of active regions on RV observations is two-fold. First, as dark spots rotate on the stellar surface, they distort the profile of spectral lines, removing a small contribution first from the red wing of each spectral line, then from the line core, and then from the blue wing. The second effect is more subtle: in the absence of active regions, spectral lines of Sun-like stars display a net blue-shift due to granulation: the emission from the hot, up-welling material in the granules, which is blue-shifted, dominates over that from the cooler material falling back down in the inter-granular lanes). In a facula, i.e.\ a region of enhanced magnetic flux density compared to the ``clean'' photosphere, convection is suppressed, leading to a localised reduction in this ``convective blue-shift''. Faculae have very small photometric contrast but can cover a much larger area than dark spots, so their RV signature can dominate over that of spots, especially for moderately slow rotators like the Sun with modest magnetic fields \citep{2010A&A...512A..39M}.

@@

\begin{figure}[ht]
  \centering
  \script{scaling.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/scaling.pdf}
  \caption{How the computational cost of a single evaluation of the GP likelihood function (\autoref{eq:gp_like}) scales with the size of the dataset using different algorithms.
  In each case, the kernel function is a one-dimensional Mat\'ern-3/2 with randomly generated data.
  The top two lines labelled ``direct (CPU)'' and ``direct (GPU)'' show the cost of directly evaluating \autoref{eq:gp_like} directly using the Cholesky factorization of $\mathbf{K}$ provided by a tuned and optimised linear algebra library.
  The CPU solver was run using a single-threaded implementation, and the GPU version was executed on an \textsf{A100 NVIDIA} GPU.
  In both cases, the asymptotic scaling of the algorithm is approximately $\mathcal{O}(N^3)$ for $N$ data points.
  The three lines labelled ``celerite ($J$ terms)'' show the performance of the \project{celerite} algorithm \citep[][as implemented by the \project{tinygp} package]{celerite} for a sum of $J$ Mat\'ern-3/2 kernels, since the computational cost of that method scales with the ``complexity'' of the kernel.
  The asymptotic performance of each of these curves is linear in the number of data points, $\mathcal{O}(N)$.}
  \label{fig:scaling}
\end{figure}

@@

Alongside this review, we have released all the source code to generate the figures in this document using the \project{tinygp} library\footnote{\url{https://tinygp.readthedocs.io}} \citep{tinygp}, which is a relatively new open source \project{Python} library for GP modelling.
While this library hasn't yet been widely used in the astrophysics literature, we believe\footnote{Note that the development of the \project{tinygp} library is led by D.\ Foreman-Mackey, an author of this review.} that it is a good choice for many applications.
\project{tinygp} uses the \project{JAX}\footnote{\url{https://jax.readthedocs.io}} library as its computational backend, giving it good computational performance (including support for GPU hardware) with a pure-\project{Python} user interface.
\project{JAX} also provides support for automatic differentiation \citep[e.g.,][]{autodiff}, which can be used to accelerate all the inference schemes described in Sections~\ref{sec:inference} and~\ref{sec:model_selection}.

@@

After a very brief historical introduction, we started this review by introducing two simulated examples, which demonstrate the two main classes of use for GPR in astronomy, namely to account for a nuisance signal (in the case of a planetary transit) or to model a stochastic signal of interest (in the quasar time-delay example). The source code for these and other worked examples in the review is available alongside this review in the form of \project{Jupyter} notebooks that can be accessed by clicking on the icon next to each figure caption, and interested users are encouraged to use these to further their understanding of the methodology and potentially adapt the code to their own datasets. Next, we gave a succinct introduction to the key equations underlying GPR regression, and sketched out a typical workflow for applying GPR to a new dataset. Like any data analysis method, GPR requires the user to make a number of choices, the most important of which is the choice of covariance function. We listed some of the most widely used covariance functions, outlined how to construct more complex ones, and gave practical advice on how to assess the validity of one's modelling choices. These strategies include standard Bayesian model comparison but also cross-validation, which can be more appropriate in cases where predictive power is important.

@@

SA gave a set of lectures on GPR for time-domain astronomy as part of the 2021 \emph{Saas Fee Advanced Course in Astrophysics}, for which PDF lecture notes are available at \url{shorturl.at/hrstz}, along with {\sc Python} exercises in the form of \project{Jupyter} notebooks.
% \suz{This simple URL reference will need to be replaced with a citation to the official lecture notes published by Springer if those are published before the present review.}
